Estimating duration using survival analysis

### 2023-06-09 cis-perfect-testing.tex
Notes on changes needed
Edit perf chapter

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
We make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.

I utilise survival analysis for analysing this data.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, \ie becoming detectable.
The \emph{terminating event} is when the duration starts, \ie becoming no longer detectable.

@@

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \cref{perf-test:fig:double-interval-censor}).
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who transitions negative at time 7 to positive at time 14, bounding the start of the episode; similarly, the transition positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

Interval censoring is a well-studied area in the biostatistics and broader statistical literature\todo{interval censoring refs}.
Double interval censoring is less frequent, although has been considered previously\todo{double interval censoring refs}.

@@

Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.

@@

\section{Modelling the duration}\label{perf-test:sec:model}
Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
Only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

\subsection{Independent priors}

A common assumption throughout Bayesian analyses is independence of the parameters in the prior.
This is convenient because we can then use univariate distributions.
\todo{This feels a bit clumsy/patronising to the reader. Maybe change the phrasing?}
A Beta distribution for each hazard parameter, $\lambda_t$, is a natural choice due to it being a probability.

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-prior}),
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The first step, approximating the \cref{ATACCC} posterior as a multivariate normal, is done through the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).

@@

\subsection{Smoothing priors}

A final possible prior are ones to favour smoothness in the hazard.
This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting.
Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
A smoothing prior encodes this information within the analysis.

@@

To form the duration distribution used in the simulation, we combine these two estimates.
The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible). ERR... I'M NOT DOING THIS ANYMORE OH NO!!
Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
Define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.

@@

Discussion points:
\begin{enumerate}
  \item Other priors (GPs?)
  \item Inclusion of covariates
  \item Individual vs total model
  \item Consideration of smaller sample size
\end{enumerate}

>>>>

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.

I utilise survival analysis for analysing this data.
Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, \ie becoming detectable.
The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.

@@

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{intro:sec:cis}).
\todo{ref to description of CIS}
Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
A more familiar setting for this work

@@

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{imperf-test}).

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

\subsection{Independent priors}
Take the standard assumption of independent priors on each parameter.
A Beta distribution is then a natural choice for each hazard parameter, $\lambda_t$, as they are probabilities.

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{ATACCC} and the meta-analysis of \textcite{cevikShedding}.
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The first step, approximating the \cref{ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).

@@

\subsection{Smoothing priors}
The final class of priors I consider are ones that aim to favour smoothness in the hazard.
This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting (\eg: \autocite{bacchettiNonparametric}).
Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
A smoothing prior encodes this information within the analysis.

@@

To form the duration distribution used in the simulation, we combine these two estimates.
The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible). ERR... I'M NOT DOING THIS ANYMORE OH NO!!
Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.

@@

Discussion points:
\begin{enumerate}
  \item Other priors (GPs?)
  \item Inclusion of covariates: age, previous exposure status, variant.
  \item Individual vs total model
  \item Consideration of smaller sample size
\end{enumerate}
